### 2011-02-03 docs/architecture/architecture.tex
lots of reorganization/clean-ups

@@

In this section we describe some of the standard problems that come
with MAESTRO, how to run the code, some basic runtime parameters, and
how to look at the output.

@@

As the code runs, it will output both plotfiles and checkpoints.
By default, the plotfiles will be named {\tt plt}{\em nnnnn}, where
the number {\em nnnnn} is the timestep number when the file was
outputted.  Similarly, the checkpoints are named {\tt chk}{\em
  nnnnn}.  BoxLib plotfiles and checkpoints are actually directories,
with the data stored in subdirectories grouped by refinement level.
Details of the simulation (number of processors used, output date,
output directory, runtime parameter values, ...) are stored in
the {\tt job\_info} file in each plotfile directory.

@@

\item {\tt max\_levs } is the maximum number of grid levels in the AMR
  hierarchy to us.  {\tt max\_levs = 1} indicates running with only a
  single level spanning the whole domain.

@@

@@

\subsection{Refinement Criteria}

@@

A basic {\tt GNUmakefile} begins with:
\begin{verbatim}
  NDEBUG := t
  MPI    :=
  OMP    :=
\end{verbatim}
Here, {\tt NDEBUG} is true is we are building an optimized executable.
Otherwise, the debug version is built---this typically uses less
optimization and adds various runtime checks through compiler flags.
{\tt MPI} and {\tt OMP} are set to true if we want to use either MPI
or OpenMP for parallelization.

The MAESTRO build system knows what options to use for various
compiler families.  The {\tt COMP} flag specifies which compiler to
us.  Allowed values include {\tt Intel}, {\tt gfortran}, {\tt PGI},
and {\tt PathScale}.  The specific details of these choices are
defined in the {\tt MAESTRO/mk/} directory.
\begin{verbatim}
  COMP := Intel
  MKVERBOSE := t
\end{verbatim}
{\tt MKVERBOSE} set to true will echo the build commands to the
terminal as the are executed.

The next two lines define where the top of the source tree is
include the basic Makefile information (like compiler
flags).
\begin{verbatim}
  FPARALLEL := ../..

  include $(FPARALLEL)/mk/GMakedefs.mak
\end{verbatim}

A MAESTRO application is built from several pacakges (the
multigrid solver, an EOS, a reaction network, etc.).  The core
MAESTRO packages are always included, so a problem only needs
to define the EOS, reaction network, and conductivities to
use, as well as any extra, problem specific files.  
\begin{verbatim}
EOS_DIR := extern/EOS/helmeos   
CONDUCTIVITY_DIR := extern/EOS/conducteos 
NETWORK_DIR := extern/networks/ignition_chamulak

EXTRA_DIR := extern/random
\end{verbatim}
Generally, one does not need to include the problem directory itself
in {\tt EXTRA\_DIR}, unless there are unique source files found there,
described in a {\tt GPackage.mak} file.  These variables are
interpreted by the {\tt GMaestro.mak} file and used to build a master
list of packages called {\tt Fmdirs}.  The build system will attempt
to build all of the files listed in the various {\tt GPackage.mak}
files found in the {\tt Fmdirs} directories.  Furthermore, {\tt
  Fmdirs} will be will be added to the {\tt make} {\tt VPATH}, which
is the list of directories to search for source files.  The problem
directory will always be put first in the {\tt VPATH}, so any source
files placed there override those with the same name found elsewhere
in the source tree.  Some packages (for instance, the {\tt helmeos}
EOS) require Fortran include files.  The {\tt Fmincludes} variable
lists all those directories that contain include files that are
inserted into the Fortran source at compile time via the {\tt include}
statement.  Presently, the only instance of this is with the Helmholtz
general equation of state found in {\tt extern/EOS/helmeos/}.  This is
automatically handled by the {\tt GMaestro.mak} instructions.

Runtime parameters listed in the {\tt MAESTRO/\_parameters} file are
parsed at compile time and the file {\tt probin.f90} is written and
compiled.  This is a Fortran module that holds the values of the
runtime parameters and makes them available to any routine.  The
variable {\tt PROBIN\_PARAMETERS} should list all of the files that
contain runtime parameters for the current problem.  You must always
include the main {\tt ../\_parameters} file (i.e.\ {\tt
  MAESTRO/\_parameters}).  Sometimes a problem defines their own
runtime parameters (see \S~\ref{sec:def_runtime_param}), in which
case, its parameter file should also appear in {\tt
  PROBIN\_PARAMETERS}.
\begin{verbatim}
  PROBIN_PARAMETERS := ../_parameters
\end{verbatim}

The final line in the {\tt GNUmakefile} include the rules to actually
build the executable.
\begin{verbatim}
  include $(FPARALLEL)/MAESTRO/GMaestro.mak
\end{verbatim}

As mentioned above, any source files placed in the problem directory
override a files with the same name found elsewhere in the source
tree.  This allows you to create a problem-specific version of any
routine.  Source files that are unique to this problem (i.e.\ there is
no file with the same name elsewhere in the source tree) need to be
listed in a file {\tt GPackage.mak} in the problem directory.

@@

MAESTRO initial models are read in by the {\tt init\_base\_state}
routine.  This expects the initial model to contain a header giving
the number of variables and their names, followed by rows of data
giving the coordinate and data values at that coordinate.  The initial
model should contain the same species data (as mass fractions) as
defined in the {\tt network} module used by the MAESTRO problem.

@@

>>>>

\section{MAESTRO Basics}

MAESTRO models problems that are in tight hydrostatic equilibrium.
The fluid state is decomposed into a 1-d radial base state that
describes the hydrostatic structure of the star or atmosphere, and a
2- or 3-d Cartesian full state, that captures the departures from
hydrostatic equilibrium.  Two basic geometries are allowed.  A {\em plane-parallel}
geometry assumes that the domain is thin compared to the radius of curvature
of the star, and therefore the 1-d base state is perfectly aligned with
the Cartesian state.  A {\em spherical} geometry is for modeling an entire star.  Here,
the 1-d base state is not aligned with the Cartesian state.  Figure~\ref{fig:base_state}
shows this geometry.

\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{base_spherical}
\caption{\label{Fig:base_state} (Left) For multi-level problems
  in planar geometry, we force a direct alignment between the radial
  array cell centers and the Cartesian grid cell centers by allowing
  the radial base state spacing to change with space and time.
  (Right) For multi-level problems in spherical geometry, since there
  is no direct alignment between the radial array cell centers and
  the Cartesian grid cell centers, we choose to fix the radial base
  state spacing across levels. Figure taken from~\cite{multilevel}}
\end{figure}

MAESTRO can use adaptive mesh refinement to focus resolution on
complex regions of flow.  For Cartesian/plane-parallel geometries, all
cells at the same height must be at the same level of refinement.
This restriction is to allow for the base state to directly align with
the Cartesian state everywhere.  For spherical geometries, there is no
such restriction (again, see Figure~\ref{Fig:base_state}).

@@

In this section we give an overview of MAESTRO, including some of the
standard problems, how to run the code, some basic runtime parameters,
and how to look at the output.

@@

As the code runs, it will output both plotfiles and checkpoints as
well as one of more text diagnostic files (like {\tt
  maestro\_diag.out}) with integral or extrema information (like
maximum Mach number) from each timestep.  By default, the plotfiles
will be named {\tt plt}{\em nnnnn}, where the number {\em nnnnn} is
the timestep number when the file was outputted.  Similarly, the
checkpoints are named {\tt chk}{\em nnnnn}.  BoxLib plotfiles and
checkpoints are actually directories, with the data stored in
subdirectories grouped by refinement level.  Details of the simulation
(build information, number of processors used, output date, output
directory, runtime parameter values, ...) are stored in the {\tt
  job\_info} file in each plotfile directory.

@@

\item {\tt max\_levs } is the maximum number of grid levels in the AMR
  hierarchy to use.  {\tt max\_levs = 1} indicates running with only a
  single level spanning the whole domain.

@@

  For multilevel problems, the parameter {\tt max\_grid\_size\_1}
  controls the maximum extent on level 1 (the base grid), {\tt
    max\_grid\_size\_2} controls the maximum extent on level 2, and
  {\tt max\_grid\_size\_3} controls the maximum extent on levels 3 and
  higher.

@@

@@

A basic {\tt GNUmakefile} begins with:
\begin{lstlisting}[language=bash,mathescape=false]
  NDEBUG := t
  MPI    :=
  OMP    :=
\end{lstlisting}
Here, {\tt NDEBUG} is true if we are building an optimized executable.
Otherwise, the debug version is built---this typically uses less
optimization and adds various runtime checks through compiler flags.
{\tt MPI} and {\tt OMP} are set to true if we want to use either MPI
or OpenMP for parallelization.  If {\tt MPI := t}, you will need to
have the MPI libraries installed, and their location may need to be 
specified in {\tt MAESTRO/mk/GMakeMPI.mak}.

The next line sets the compiler to be used for compilation:
\begin{lstlisting}[language=bash,mathescape=false]
  COMP := Intel
\end{lstlisting}
The MAESTRO build system knows what options to use for various
compiler families.  The {\tt COMP} flag specifies which compiler to
use.  Allowed values include {\tt Intel}, {\tt gfortran}, {\tt PGI},
and {\tt PathScale}.  The specific details of these choices are
defined in the {\tt MAESTRO/mk/comps/} directory.

{\tt MKVERBOSE} set to true will echo the build commands to the
terminal as the are executed.
\begin{lstlisting}[language=bash,mathescape=false]
  MKVERBOSE := t
\end{lstlisting}

The next two lines define where the top of the source tree is
include the basic Makefile information (like compiler
flags).
\begin{lstlisting}[language=bash,mathescape=false]
  FPARALLEL := ../..

  include $(FPARALLEL)/mk/GMakedefs.mak
\end{lstlisting}
%$

A MAESTRO application is built from several pacakges (the
multigrid solver, an EOS, a reaction network, etc.).  The core
MAESTRO packages are always included, so a problem only needs
to define the EOS, reaction network, and conductivities to
use, as well as any extra, problem-specific files.  
\begin{lstlisting}[language=bash,mathescape=false]
EOS_DIR := extern/EOS/helmeos   
CONDUCTIVITY_DIR := extern/conductivity/constant
NETWORK_DIR := extern/networks/ignition

EXTRA_DIR := extern/random
\end{lstlisting}
Generally, one does not need to include the problem directory itself
in {\tt EXTRA\_DIR}, unless there are unique source files found there,
described in a {\tt GPackage.mak} file.  These variables are
interpreted by the {\tt GMaestro.mak} file and used to build a master
list of packages called {\tt Fmdirs}.  The build system will attempt
to build all of the files listed in the various {\tt GPackage.mak}
files found in the {\tt Fmdirs} directories.  Furthermore, {\tt
  Fmdirs} will be will be added to the {\tt make} {\tt VPATH}, which
is the list of directories to search for source files.  The problem
directory will always be put first in the {\tt VPATH}, so any source
files placed there override those with the same name found elsewhere
in the source tree.

Some packages (for instance, the {\tt helmeos}
EOS) require Fortran include files.  The {\tt Fmincludes} variable
lists all those directories that contain include files that are
inserted into the Fortran source at compile time via the {\tt include}
statement.  Presently, the only instance of this is with the Helmholtz
general equation of state found in {\tt extern/EOS/helmeos/}.  This is
automatically handled by the {\tt GMaestro.mak} instructions.

Runtime parameters listed in the {\tt MAESTRO/\_parameters} file are
parsed at compile time and the file {\tt probin.f90} is written and
compiled.  This is a Fortran module that holds the values of the
runtime parameters and makes them available to any routine.  The
variable {\tt PROBIN\_PARAMETERS} should list all of the files that
contain runtime parameters for the current problem.  You must always
include the main {\tt ../\_parameters} file (i.e.\ {\tt
  MAESTRO/\_parameters}).  Sometimes a problem defines their own
runtime parameters (for instance {\tt \_parameters.myproblem}), in which
case, its parameter file should also appear in {\tt
  PROBIN\_PARAMETERS} (see \S~\ref{sec:def_runtime_param}).
\begin{lstlisting}[language=bash,mathescape=false]
  PROBIN_PARAMETERS := ../_parameters
\end{lstlisting}

The final line in the {\tt GNUmakefile} includes the rules to actually
build the executable.
\begin{lstlisting}[language=bash,mathescape=false]
  include $(FPARALLEL)/MAESTRO/GMaestro.mak
\end{lstlisting}
%$

\subsubsection{Handling Problem-Specific Source Files}

As mentioned above, any source files placed in the problem directory
override a files with the same name found elsewhere in the source
tree.  This allows you to create a problem-specific version of any
routine.  Source files that are unique to this problem (i.e.\ there is
no file with the same name elsewhere in the source tree) need to be
listed in a file {\tt GPackage.mak} in the problem directory, and
the problem-directory needs to be explicitly listed in the {\tt EXTRA\_DIR}
list in the {\tt GNUmakefile}.

@@

MAESTRO initial models are read in by the {\tt extern/model\_parser}
routines.  This expects the initial model to contain a header giving
the number of variables and their names, followed by rows of data
giving the coordinate and data values at that coordinate.  The initial
model should contain the same species data (as mass fractions) as
defined in the {\tt network} module used by the MAESTRO problem.

@@

\section{Multilevel and Refinement Criteria}

\section{Particles}
